From Reversible Programs to \\ Univalent Universes and Back

### 2021-11-11 Pi/paper/discussion.tex
Fix table formatting
Co-authored-by: Jacek Karwowski <inexxt@users.noreply.github.com>
Final revisions

\begin{figure}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            \rowcolor{lightgray} $\PiLang$ & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline
            Types                          & Natural numbers               & Finite sets & 0-cells \tabularnewline
            \hline
            1-combinators                  & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
            \hline
            2-combinators                  & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \caption{Summary of the established correspondences TODO: Move above}
    \label{fig:table-summary}
\end{figure}

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms. It could
provide an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult to formalise in type theory
or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
intractable.

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse. Univalent
typoids~\cite{petrakisUnivalentTypoids2019} are another way of axiomatising univalent subuniverses.

% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which provides
an intuition on why certain constructions are hard (or impossible in the general case). There are other, different
approaches to computing with univalence,
in~\cite{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in Cubical
Type Theory\cite{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

%%%%%%%%%%%%%%
\paragraph{Free Symmetric Monoidal Category} The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, Petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005}, and
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}. The forgetful functor
from $\SymMonCat$, the 2-category of (small) symmetric monoidal categories, strong symmetric monoidal functors, and
monoidal natural transformations, to the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal
category $\FSM[\CCat{C}]$ on a category $\CCat{C}$. This is a 2-monad on
$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. Its
construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.  Concretely, the objects of
$\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$;
morphisms between $(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
$C^{n}$ by action. The groupoid $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.

>>>>

\begin{table}
    \caption{Summary of the established correspondences}
    \begin{center}
        \begin{tabular}{cccc}
            \hline
            $\PiLang$     & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
            \hline\hline
            Types         & Natural numbers               & Finite sets & 0-cells \tabularnewline
            1-combinators & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
            2-combinators & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
            \hline
        \end{tabular}
    \end{center}
    \label{fig:table-summary}
\end{table}

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
    {homotopy.io}, Agda, Coq and Lean. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms, which could
provide an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult to formalise in type theory
or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
intractable.

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ is constructed using path induction. The term $\term{ua} :
{A \eqv B \to A \id_{\UU} B}$, its computation rule $\term{ua_\beta} : {(e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e))
\id e}$, and its extensionality rule $\term{ua_\eta} : {(p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})}$ are
generally added as postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full
univalence axiom ${(A \eqv B) \eqv (A \id_{\UU} B)}$ \cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse. Univalent
typoids~\cite{petrakisUnivalentTypoids2019} are another way of axiomatising univalent subuniverses.

% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new way of thinking about computing using the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in Cubical
Type Theory\cite{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

%%%%%%%%%%%%%%
\paragraph{Free Symmetric Monoidal Category} The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, Petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005}, and
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}. The forgetful functor
from $\SymMonCat$, the 2-category of (small) symmetric monoidal categories, strong symmetric monoidal functors, and
monoidal natural transformations, to the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal
category $\FSM[\CCat{C}]$ on a category $\CCat{C}$. This is a 2-monad on
$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. A
construction of $\FSM$ is given by~\citet{abramskyAbstractScalarsLoops2005}. Concretely, the objects of $\FSM[\CCat{C}]$
are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$; morphisms between
$(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and $\lambda_{i} : A_{i} \to
B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction $\int F$ of the functor $F
: \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning each natural number $n$ to
the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on $C^{n}$ by action. The groupoid
$\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.

### 2021-11-11 Pi/paper/equivalence.tex
Final revisions

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
$\PiHatCat$, have 0-cells for types ($U$, $\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and
$\isoh$), and 2-cells for 2-combinators ($\Iso$, $\Isop$, and $\Isoh$). We can show that these syntactic categories here
are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict, or
locally-posetal, because there is at most one 2-cell between compatible 1-cells.

@@

As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure.~\footnote{Since each object
is a natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.}

To produce a braiding $n + m \isoh m + n$ from adjacent tranpositions, we recursively traverse the left subexpression,
swapping each element using adjacent transpositions along the elements on the right, placing it in the right position.
The computational content of this translation can be visualised using tree transformations, for the recursive case
in~\cref{fig:plusplusswap}. The challenging part is showing that these moves are coherent with respect to 2-combinators.

@@

\begin{definition}[$\evalh$]
  \begin{align*}
    \evalh_{0} & : \UHat \to \UFin                                                          \\
    \evalh_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
    \evalh_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalh_{1}(c_{1}) \id \evalh_{1}(c_{2})
  \end{align*}
\end{definition}

>>>>

To state our results formally, we organise the syntax for each language using its syntactic category. We define them
formally in the appendix, and only state our results here. For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$
languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and $\PiHatCat$, have 0-cells for types ($U$,
$\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and $\isoh$), and 2-cells for 2-combinators ($\Iso$,
$\Isop$, and $\Isoh$). These syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and
2-cells are invertible. They are also locally-strict, or locally-posetal, because there is at most one 2-cell between
compatible 1-cells.

@@

As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure. Since each object is a
natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.

To produce a braiding $n + m \isoh m + n$ from adjacent transpositions, we recursively traverse the left subexpression,
swapping each element using adjacent transpositions along the elements on the right, placing it in the correct position.
The computational content of this translation can be visualised using tree transformations, for the recursive case
in~\cref{fig:plusplusswap}. The challenging part is showing that these moves are coherent with respect to 2-combinators.

@@

\begin{definition}[$\evalh$]
  \label{def:evalh}
  \begin{align*}
    \evalh_{0} & : \UHat \to \UFin                                                          \\
    \evalh_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
    \evalh_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalh_{1}(c_{1}) \id \evalh_{1}(c_{2})
  \end{align*}
\end{definition}

### 2021-11-11 Pi/paper/examples.tex
Final revisions

\begin{figure}
  \begin{tabular}{c@{\qquad}c}
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        ccx q[1], q[2], q[0];
        cx  q[1], q[0];
        cx  q[2], q[0];
  \end{BVerbatim}
      \inkscape[0.5]{qis1}
    \end{minipage}
     &
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        cx  q[1], q[0];
        x   q[1];
        ccx q[1], q[2], q[0];
        x   q[1];
  \end{BVerbatim}
      \inkscape[0.5]{qis2}
    \end{minipage}
  \end{tabular}
  \caption{$\mathit{reversibleOr}$ in Qiskit}
  \label{fig:qiskit}
\end{figure}

There is a wealth of manual and algorithmic approaches for producing circuits such as these
two~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left was manually produced using a standard synthesis
algorithm for reversible circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced using an approach
that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
more than two inputs).

@@

Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) =
a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible
specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
disjunction~$\underline{\vee}$ is applied bitwise. When $h=0$ the second component of the result of $f^R$ is the result
of the original modular exponentiation $f$.
%%
%% the specification of the circuit is relatively straightforward to
%% calculate. Here it is for $a=11$ and $N=15$:
%% \[\begin{array}{rcll}
%% g(r,h) &=& \left\{ \begin{array}{ll}
%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
%%                                 \end{array}\right.
%% \end{array}\]
%%
However, as explained in standard accounts of the algorithm (e.g., the
Qiskit implementation), producing an efficient modular exponentiation circuit
from this specification is not straightforward and is actually the bottleneck in
Shor’s algorithm. Typical derivations of the circuit start from elementary
gates, build a circuit for reversible disjunction (like the two circuits above),
reversible conjunction, a circuit for a half-adder, a circuit for computing the
carry, progressing to a circuit for modular addition, which is used to build a
circuit for modular multiplication, and then finally a circuit for modular
exponentiation taking care at each step to avoid the exponential blowup (e.g.,
by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.

@@

% From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result. The $\mathit{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):

>>>>

\begin{figure}
  \begin{tabular}{c@{\qquad}c}
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        ccx q[1], q[2], q[0];
        cx  q[1], q[0];
        cx  q[2], q[0];
  \end{BVerbatim}
      \inkscape[0.5]{qis1}
    \end{minipage}
     &
    \begin{minipage}[t]{0.45\linewidth}
      \begin{BVerbatim}[fontsize=\small]
        cx  q[1], q[0];
        x   q[1];
        ccx q[1], q[2], q[0];
        x   q[1];
  \end{BVerbatim}
      \inkscape[0.5]{qis2}
    \end{minipage}
  \end{tabular}
  \caption{$\mathsf{reversibleOr}$ in Qiskit}
  \label{fig:qiskit}
\end{figure}

There is a wealth of manual and algorithmic approaches for producing circuits such as these
two~\cite{maslovReversibleLogicSynthesis2003,shendeSynthesisReversibleLogic2003}. The circuit on the left was manually
produced using a standard synthesis algorithm for reversible circuits~\cite{millerTransformationBasedAlgorithm2003}. The
circuit on the right was produced using an approach that analyzes the recursive structure of $\mathsf{reversibleOr}$
(and would generalise to computing the disjunction of more than two inputs).

@@

Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) =
a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{toffoliReversibleComputing1980},
a reversible specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
disjunction~$\underline{\vee}$ is applied bitwise. When $h=0$ the second component of the result of $f^R$ is the result
of the original modular exponentiation $f$.
%%
%% the specification of the circuit is relatively straightforward to
%% calculate. Here it is for $a=11$ and $N=15$:
%% \[\begin{array}{rcll}
%% g(r,h) &=& \left\{ \begin{array}{ll}
%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
%%                                 \end{array}\right.
%% \end{array}\]
%%
However, as explained in standard accounts of the algorithm (e.g., the Qiskit implementation), producing an efficient
modular exponentiation circuit from this specification is not straightforward and is actually the bottleneck in Shor’s
algorithm. Typical derivations of the circuit start from elementary gates, build a circuit for reversible disjunction
(like the two circuits above), reversible conjunction, a circuit for a half-adder, a circuit for computing the carry,
progressing to a circuit for modular addition, which is used to build a circuit for modular multiplication, and then
finally a circuit for modular exponentiation taking care at each step to avoid the exponential blowup (e.g., by
implementing exponentiation by squaring instead of repeated multiplication)~\cite{beauregardCircuitShorAlgorithm2003}.

@@

% From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result. The $\mathsf{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):

### 2021-11-11 Pi/paper/finite.tex
Final revisions

% Recap of the previous section
In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
is also known to be the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$
for every $n$. In this section, we will describe this group syntactically.

@@

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a (listed) finite set, which satisfies the invariant of
being a sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the
permutations on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate
for the generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using
generators (primitive operations) that:

\begin{itemize}
  \item swap the $i$-th element with the $(i+1)$-th element, that is, adjacent swaps, or
  \item swap the $i$-th element with the $j$-th element, for arbitrary $i$-s and $j$-s, or
  \item swap the $i$-th element with an element at a fixed position, or
  \item reverses a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
  \item cyclically shift any subset of $\Fin[n]$.
\end{itemize}

@@

A group homomorphism between groups $G$ and $H$ is a function $f : G \to H$ between the underlying sets that preserves
the group structure. Giving a group homomorphism out of the free group is equivalent to giving a function out of the
generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
category of groups and sets. We say that $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies the following
universal property.

@@

\begin{definition}[Free group]
  Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
      A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons a \cons \nil
    \sim \nil$.
\end{definition}

@@

For this to work, we first need the system to have the termination property, meaning that there should be no infinite
reductions. We observe that after throwing out reflexivity and symmetry, the right-hand sides of the relations $\cox*$
are strictly smaller than the left-hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can extract a normalisation function. This will be true if the
rewriting system is confluent -- all critical pairs, that is, terms with overlapping possible reduction rules, have to
converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges. Unfortunately, this is
not true for all critical pairs -- an example is in~\cref{fig:critical-pairs-non-converging}, where left and right
endpoints are normal with respect to the $\cox*$ relation.

@@

\noindent
Finally, we need to show that $\Sn$ and is indeed a group presentation.

@@

\noindent
Notice however, that for a group presentation, as defined in~\cref{def:presentation}, the relation needs to be on the
set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
proposition lets us lift a group structure on $\List[A]$ to a group presentation, with underlying set $\List[A+A]$.

@@

To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,
but now we will explicitly describe what these representatives look like, using an encoding called Lehmer
codes~\cite{lehmerTeachingCombinatorialTricks1960}.

>>>>

% Recap of the previous section
In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
is also the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$ for every
$n$. In this section, we will describe this group syntactically.

@@

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a (listed) finite set, which satisfies the invariant of
being a sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the
permutations on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate
for the generators of symmetric groups. For example, we could generate the symmetric group on $\Fin[n]$ by using
generators (primitive operations) that:

\begin{itemize}
  \item swap the $i$-th element with the $(i+1)$-th element, that is, adjacent swaps, or
  \item swap the $i$-th element with the $j$-th element, for arbitrary $i$-s and $j$-s, or
  \item swap the $i$-th element with an element at a fixed position, or
  \item reverse a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
  \item cyclically shift any subset of $\Fin[n]$.
\end{itemize}

@@

A group homomorphism between groups $G$ and $H$ is a function $f : G \to H$ between the underlying sets that preserves
the group structure. Giving a group homomorphism out of the free group is equivalent to giving a function out of the
generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
categories of groups and sets. We say that $F(A)$ with $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies
the following.

@@

\begin{definition}[Free group]
  Let $A$ be a set, and $\List[\blank]$ the free monoid monad. The free group $F(A)$ on $A$ is the set-quotient of
      $\List[A + A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons
      a \cons \nil \sim \nil$.
\end{definition}

@@

For this to work, we first need the system to have the termination property, meaning that there should be no infinite
reductions. We observe that, after throwing out reflexivity and symmetry, the right-hand sides of the relations $\cox*$
are strictly smaller than the left-hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
Second, we need the normal forms to be unique, so that we can extract a normalisation function. This will be true if the
rewriting system is confluent -- all critical pairs, that is, terms with overlapping possible reduction rules, have to
converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges. Unfortunately, this is
not true for all critical pairs -- an example is in~\cref{fig:critical-pairs-non-converging}, where left and right
endpoints are normal with respect to the $\cox*$ relation.

@@

\noindent
Finally, we need to show that $\Sn$ is indeed a group presentation.

@@

\noindent
Notice however, that for a group presentation, as defined in~\cref{def:presentation}, the relation needs to be on the
set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
proposition lets us lift a group structure on $\List[A]$ to a group presentation, whose underlying set is $\List[A+A]$.

@@

To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will define functions back and forth between the two
types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect to the
Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,
but now we will explicitly describe what these representatives look like, using an encoding called Lehmer
codes~\cite{lehmerTeachingCombinatorialTricks1960}.

### 2021-11-11 Pi/paper/introduction.tex
Final revisions

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronsonClassificationReversibleBit2017}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic data types, as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012, jamesTheseusHighLevelLanguage2014}. The
syntax of $\PiLang$ is inspired by the sound and complete axiomatisation of type isomorphisms of lambda
calculi~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the bicartesian structure, that is, product and sum
types. $\PiLang$ programs correspond to type isomorphisms of finite types, and our observation is that the syntax of
$\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).

@@

Using this algorithm, we can turn a type isomorphism into a permutation of finite sets, relating the operational
semantics of $\PiLang$ as permutations of finite sets of bits, to our denotational semantics. This allows us to
establish full abstraction and adequacy.

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
state our result as a Curry-Howard-Lambek correspondence for Reversible
Logic~\cite{sparksSuperstructuralReversibleLogic2014}, Reversible Programming
Languages~\cite{jamesInformationEffects2012}, and Symmetric Rig Groupoids~\cite{laplazaCoherenceDistributivity1972,
kellyCoherenceTheoremsLax1974, caretteComputingSemiringsWeak2016}, using which we can build a toolbox of technical
devices for reasoning about reversible circuits.

>>>>

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronsonClassificationReversibleBit2017}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic data types, as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012, jamesTheseusHighLevelLanguage2014}. The
syntax of $\PiLang$ is inspired by the sound and complete axiomatisation of type isomorphisms of lambda
calculi~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the bicartesian structure, that is, product and sum
types. $\PiLang$ programs correspond to type isomorphisms of finite types, and in this paper, we make the observation
that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).

@@

Using this algorithm, we can turn a type isomorphism into a bijection of finite sets, relating the operational semantics
of $\PiLang$ as permutations of finite sets of bits, to our denotational semantics. This allows us to establish full
abstraction and adequacy.

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets, bijections, and homotopies, formalised in
HoTT/UF. We state our result as a Curry-Howard-Lambek correspondence for Reversible
Logic~\cite{sparksSuperstructuralReversibleLogic2014}, Reversible Programming
Languages~\cite{jamesInformationEffects2012}, and Symmetric Rig Groupoids~\cite{laplazaCoherenceDistributivity1972,
kellyCoherenceTheoremsLax1974, caretteComputingSemiringsWeak2016}, using which we can build a toolbox of technical
devices for reasoning about reversible circuits.
